### 2014-12-12 text/Chapter03.tex
Proofreading and English conventions (Chapter 4)

We look at resources in distributed programming in more details in Section~\ref{sec:applications-flat-distr}.
The important point here is that in flat coeffect systems, contextual requirements are 
\emph{split} between the call-site and declaration-site. Furthermore, there is no syntactic
structure that determines how the requirements are split. In the case of distributed
programming, the resources can be freely associated with either of the two sites.

@@

Figure~\ref{fig:applications-flat-impl} shows a type system that tracks the set of expression's 
implicit parameters. The type system uses judgements of the form $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$
meaning that an expression $e$ has a type $\tau$ in a free-variable context $\Gamma$ with a set 
of implicit parameters specified by $\cclrd{r}$. The annotations $\cclrd{r},\cclrd{s},\cclrd{t}$ are 
finite partial functions mapping implicit parameter names to types, \ie~$\cclrd{r},\cclrd{s},\cclrd{t} \subseteq 
\ident{Names} \mapsto \ident{Types}$. The expressions include \ident{?param} to read implicit
parameter and \kvd{letdyn} to bind an implicit parameter. The types are standard, but functions are
annotated with the set of implicit parameters that must be available on the call-site, \ie~
$\tau_1 \xrightarrow{\cclrd{s}} \tau_2$.

@@

When we read the rules from the top to the bottom, application (\emph{app}) and let binding 
(\emph{let}) simply union the context requirements of the sub-expressions. However, lambda abstraction
(\emph{abs}) is where the example differs from effect systems. The implicit parameters required by
the body $\cclrd{r} \cup \cclrd{s}$ can be freely split between the declaration-site ($\coctx{\Gamma}{\cclrd{r}}$)
and the call-site ($\tau_1 \xrightarrow{\cclrd{s}} \tau_2$). Finally, (\emph{letdyn}) defines an implicit
parameter and removes it from the set of requirements.

@@

The next typing derivation ($3$) requires the parameter \ident{?a} from both the declaration-site and 
the call-site. This means that, at runtime, two values will be available. Our semantics for the 
system describes \emph{dynamic rebinding}, meaning that when the caller provides a value for a 
parameter that is already specified by the declaration-site, the new value hides the old one. This
means that only the value from the call-site is actually used. This ($4$) gives a more precise
typing for this situation.

@@

The most interesting rules are (\emph{abs}) and (\emph{app}). In abstraction, we get two dictionaries
$f$ and $g$ (from the declaration-site and call-site, respectively), which are combined and passed 
to the body of the function. The semantics prefers values from the call-site, which is captured by
the $\uplus$ operation. In application, we first evaluate the expression $e_1$, then $e_2$ and finally
call the returned function. The three calls use (possibly overlapping) restrictions of the dictionary
as required by the static types.

@@

When using the usualmonadic semantics based on the reader monad, the semantics of the (\emph{abs})
rule would be modified as follows:
%
\begin{equation*}
\hspace{-0.5em}
\begin{array}{l}
  \sem{\coctx{\Gamma}{\cclrd{\emptyset}} \vdash \lambda y. e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } = 
      \lambda ((x_1, \ldots, x_n), \_) \rightarrow\\[-0.25em]
  \qquad\lambda (y, g) \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\cclrd{s}} \vdash e : \tau_2 }~((x_1, \ldots, x_n, y), g)      
\end{array}
\end{equation*}
%
Note that the declaration-site dictionary
is ignored and the body is called with only the dictionary provided by the call-site. This is
a consequence of the fact that monadic functions are always pure values created using monadic
\emph{unit}, which turns a function $\tau_1 \rightarrow \mtyp{\cclrd{r}}{\tau_2}$ into a monadic
computation with no side-effects $\mtyp{\cclrd{\emptyset}}{\tau_1 \rightarrow \mtyp{\cclrd{r}}{\tau_2}}$.

@@

\paragraph{Summary.}
Implicit parameters are the simplest example of a system where function abstraction does not 
``delay'' all impurities of the body. Here, the term ``delay'' refers to the fact that some 
implicit parameters may be captured (from the declaration-site) at the time when the function is 
defined, but before it is executed. As discussed in Section~\ref{sec:applications-structure-lam},
this is the defining feature of \emph{coeffect} systems.

In this section, we have seen how this affects both the type system and the semantics of the 
language. In the type system, the (\emph{abs}) rule places context-requirements on both the 
declaration-site and the call-site. For implicit parameters, this rule introduces non-determinism
in the type-inference, because the parameters can be split arbitrarily. However, as we show in the 
next section, this is not always the case. Semantically, lambda abstraction \emph{merges} two 
parts of context (implicit parameter dictionaries) that are provided by the call-site and 
declaration-site.

@@

The type system and semantics for rebindable resources are essentially the same as those for
implicit parameters. Primitive requirements are introduced by the $\kvd{access}$ keyword. 
Lambda abstraction splits the requirements between declaration-site (capturing remote reference) 
and call-site (representing rebinding). For this reason, we do not discuss the system in details
and instead look at other uses.

@@

Sub-coeffecting and application are trivially dual to the rules for implicit parameters. We just track
supported environments using intersection as opposed to tracking required parameters using union.
However, this symmetry does not hold for lambda abstraction (\emph{abs}), which still uses \emph{union}.
This models the case when there are two ways of executing the function:
%
\begin{itemize}
\item[--] The function is represented as executable code for an call-site environment
  and is executed there, possibly after it is marshalled and transferred to another machine.
\vspace{-0.5em}
\item[--] The function body is compiled for the declaration-site environment; the value
  that is returned is a remote reference to the code and function calls are performed as remote invocations.
\end{itemize}
%
This example ignores important considerations -- for example, it is likely desirable to make this
difference explicit (\eg~using explicit wrapping of unevaluated expressions) and the implementation
also needs to be clarified. For a system that does this, see \eg~ML5 \cite{app-distributed-ml5}).
The key point of our brief example is that the algebraic structure of coeffect annotations may be more 
complex and use, for example, $\cap$ for application and $\cup$ for abstraction.

The second system, shown in Figure~\ref{fig:applications-flat-cross} (b) is inspired by the API
level requirements in Android. Coeffect annotations are simply numbers representing the level
($\cclrd{r}, \cclrd{s}, \cclrd{t} \in \mathbb{N}$). Levels are ordered increasingly, so we can
always require higher level (\emph{sub}). The requirement on function application (\emph{app}) is
the highest level of the levels required by the sub-expressions and the function. The system uses 
yet another variant of lambda abstraction (\emph{abs}) -- the requirements of the body are duplicated
and placed on \emph{both} the declaration-site and the call-site.

@@

\paragraph{Flat liveness analysis.}
In this section, we discuss a restricted form of liveness analysis. We do not track liveness of 
\emph{individual} variables, but of the \emph{entire} variable context. This is not practically
useful, but it provides an interesting insight into how flat coeffects work. A per-variable liveness
analysis can be captured using structural coeffects and is discussed in Section~\ref{sec:applications-struct-live}.
Consider the following two examples:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{constant42} = \lambda \ident{x} \rightarrow 42\\
\kvd{let}~\ident{constant} = \lambda \ident{value} \rightarrow \lambda \ident{x} \rightarrow \ident{value}
\end{array}
\end{equation*}
%
The body of the first function is just a constant $42$ and so the context of the body is marked
as \emph{dead}. The parameter (call-site) of the function is not used and can also be marked as dead. 
Similarly, no variables from the declaration-site are used and so they are also marked as dead.

In contrast, the body of the second function accesses a variable \ident{value} and so the body 
of the function is marked as \emph{live}. In the flat system, we do not track \emph{which} 
variable was used and so we have to mark both the call-site and the declaration-site as live (this will
be refined in a structural version of the system).

@@

The (\emph{abs}) rule duplicates the annotation of the body, similarly to the cross-compilation 
example in Figure~\ref{fig:applications-flat-cross}. When the body accesses any variables, it 
requires both the argument and the variables from declaration-site. When it does not use any variables,
it marks both as dead. Finally, the (\emph{let}) rule annotates the composed expression with the
liveness of the expression $e_2$ -- if the context of $e_2$ is live, then it also requires variables
from $\Gamma$; if it is dead, then it does not require $\Gamma$ or $x$. 
As further discussed later in Section~?, the (\emph{let}) rule is again just a syntactic sugar for 
$(\lambda x.e_2)~e_1$. This follows from the simple observation that 
$\cclrd{r} \sqcup (\cclrd{s} \sqcap \cclrd{r}) = \cclrd{r}$.

@@

\paragraph{Examples.}
Before looking at the semantics, we consider a number of simple examples to demonstrate the
key aspects of the system. Full typing derivations are shown in Appendix~\ref{sec:appendixa-liveness}:
%
\begin{equation*}
\begin{array}{lll}
(\lambda x . 42)~y &~\hspace{1em}~&(1)\\
\ident{twoTimes}~42          &&(2)\\
(\lambda x . x)~42 &&(3)\\
\end{array}
\end{equation*}
%
In the first case, the context is dead. In ($1$), the function's parameter is dead and so the
overall context is dead, even though the argument uses a variable $y$ -- the semantics evaluates 
the function without passing it an actual argument. In the second case ($2$), the function is
a variable that needs to be obtained and so the context is live. In the last case ($3$), the 
function accesses a variable and so its declaration-site is marked as requiring the context 
(\emph{abs}). This is where structural coeffect analysis would be more precise -- the system shown
here cannot capture the fact that $x$ is a bound variable.

@@

\paragraph{Summary.}
Unlike with implicit parameters, lambda abstraction for liveness analysis does not introduce 
non-determinism. It simply duplicates the context requirements. However, this still matches the
property of coeffects that impurities cannot be delayed or thunked and attached just to the 
function arrow -- we place requirements on both call-site and declaration-site.

@@

In function abstraction (\emph{abs}), the requirements of the body are duplicated on the declaration-site
and the call-site as in liveness analysis. If the body requires $\cclrd{n}$ past values, it may access
$\cclrd{n}$ values of any variables -- including those available in $\Gamma$, as well as the parameter
$x$. Finally, the (\emph{let}) rule simply adds the two requirements. This corresponds to the sequential
composition operation, but it is also a rule that we obtain by treating let-binding as a syntactic 
sugar for $(\lambda x.e_2)~e_1$.

@@

Lambda abstractions (\emph{abs}) receives two lists of the same size -- one containing values of
the variables (list of products) from the declaration-site $\langle \mathbf{v}_0, \ldots, \mathbf{v}_{\cclrd{n}} \rangle$ 
and one containing the argument (list of values) provided by the call-site $\langle y_0, \ldots, y_{\cclrd{n}} \rangle$.
The semantics applies the well-known \emph{zip} operation on the lists and passes the result to the
body.

@@

\paragraph{Summary.}
Despite attaching annotations to the variable context, the system for safe locking uses 
effect-style lambda abstraction. Lambda abstraction associates all requirements with the 
call-site -- a lambda function created under a lock cannot access protected memory available at 
the time of creation. It will be executed later and can only access the memory available then.
This suggests that safe locking is better seen as an effect system.

@@

\paragraph{Structural liveness.} 
Recall two examples discussed earlier where the flat liveness analysis marked the whole context
as (syntactically) live, despite the fact part of it was (semantically) dead:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{constant} = \lambda y \rightarrow \lambda x \rightarrow y\\[-0.25em]
\kvd{let}~\ident{answer} = (\lambda x \rightarrow x)~42
\end{array}
\end{equation*}
%
In the first case, the variable $x$ is dead, but was marked as live. In the second example, the
declaration-site of the \ident{answer} value is dead, but was marked as live. This is because in 
both of the expressions, \emph{some} variable is accessed. However, the (\emph{abs}) rule of flat
liveness has no way of determining \emph{which} variables are used by the body -- and, in particular,
whether the accessed variable is the \emph{bound} variable or some of the \emph{free} variables.

@@

\paragraph{Semantics.}
In the previous examples, we defined the semantics -- somewhat informally -- using a simple 
$\lambda$-calculus language to encode the model. More formally, this could be a Cartesian Closed 
Category. In that model, we can reuse variables arbitrarily and so it is not
a good fit for modelling bounded reuse. Girard et al. \cite{logic-bounded} model their bounded
linear logic in an (ordinary) linear logic where variables can be used at most once.

@@

\paragraph{Example.} As an example, consider a function $\lambda x.\kvd{prev}~(y+x)$ applied to an argument
$\kvd{prev}~(\kvd{prev}~y)$. The body of the function accesses the past value of two variables, one free
and one bound. The (\emph{abs}) rule splits the annotations between the declaration-site and call-site
of the function:
%
\begin{equation*}
\tyrule{abs}
  {\coctx{y\!:\!\mathbb{Z}, x\!:\!\mathbb{Z}}{\alift{1, 1}} \vdash \kvd{prev}~(y+x) : \mathbb{Z} }
  {\coctx{y\!:\!\mathbb{Z}}{\alift{1}} \vdash \lambda x . \kvd{prev}~(y+x) : \mathbb{Z} \xrightarrow{\cclrd{1}} \mathbb{Z} }
\end{equation*}
%
The expression always requires the previous value of $y$ and adds it to a previous value of the 
parameter $x$. Evaluating the value of the argument $\kvd{prev}~(\kvd{prev}~y)$ requires two past 
values of $y$ and so the overall requirement for the (free) variable $y$ is $3$ past values. In 
order to use the contraction rule, we rename $y$ to $y'$ in the argument:
%
\begin{equation*}
\inference
  { \coctx{y\!:\!\mathbb{Z}}{\alift{1} } \vdash \lambda x.~(\ldots) : \mathbb{Z} \xrightarrow{\cclrd{1}} \mathbb{Z} &
    \coctx{x\!:\!\mathbb{Z}}{\alift{2}} \vdash (\kvd{prev}~(\kvd{prev}~y') : \mathbb{Z} }
{\inference
  { \coctx{y\!:\!\mathbb{Z}, y'\!:\!\mathbb{Z}}{\alift{1,3}} \vdash (\lambda x.\kvd{prev}~(y+x))~(\kvd{prev}~(\kvd{prev}~y')) : \mathbb{Z} }
  { \coctx{y\!:\!\mathbb{Z}}{\alift{3}} \vdash (\lambda x.\kvd{prev}~(y+x))~(\kvd{prev}~(\kvd{prev}~y)) : \mathbb{Z} } }
\end{equation*}
%
The derivation uses (\emph{app}) to get requirements $\alift{1,3}$ and then (\emph{contr}) to take 
the maximum, showing three past values are sufficient.

>>>>

We look at resources in distributed programming in more details in Section~\ref{sec:applications-flat-distr}.
The important point here is that in flat coeffect systems, contextual requirements are 
\emph{split} between the call site and declaration site. Furthermore, there is no syntactic
structure that determines how the requirements are split. In the case of distributed
programming, the resources can be freely associated with either of the two sites.

@@

Figure~\ref{fig:applications-flat-impl} shows a type system that tracks the set of expression's 
implicit parameters. The type system uses judgements of the form $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$
meaning that an expression $e$ has a type $\tau$ in a free-variable context $\Gamma$ with a set 
of implicit parameters specified by $\cclrd{r}$. The annotations $\cclrd{r},\cclrd{s},\cclrd{t}$ are 
finite partial functions mapping implicit parameter names to types, \ie~$\cclrd{r},\cclrd{s},\cclrd{t} \subseteq 
\ident{Names} \mapsto \ident{Types}$. The expressions include \ident{?param} to read implicit
parameter and \kvd{letdyn} to bind an implicit parameter. The types are standard, but functions are
annotated with the set of implicit parameters that must be available on the call site, \ie~
$\tau_1 \xrightarrow{\cclrd{s}} \tau_2$.

@@

When we read the rules from the top to the bottom, application (\emph{app}) and let binding 
(\emph{let}) simply union the context requirements of the sub-expressions. However, lambda abstraction
(\emph{abs}) is where the example differs from effect systems. The implicit parameters required by
the body $\cclrd{r} \cup \cclrd{s}$ can be freely split between the declaration site ($\coctx{\Gamma}{\cclrd{r}}$)
and the call site ($\tau_1 \xrightarrow{\cclrd{s}} \tau_2$). Finally, (\emph{letdyn}) defines an implicit
parameter and removes it from the set of requirements.

@@

The next typing derivation ($3$) requires the parameter \ident{?a} from both the declaration site and 
the call site. This means that, at runtime, two values will be available. Our semantics for the 
system describes \emph{dynamic rebinding}, meaning that when the caller provides a value for a 
parameter that is already specified by the declaration site, the new value hides the old one. This
means that only the value from the call site is actually used. This ($4$) gives a more precise
typing for this situation.

@@

The most interesting rules are (\emph{abs}) and (\emph{app}). In abstraction, we get two dictionaries
$f$ and $g$ (from the declaration site and call site, respectively), which are combined and passed 
to the body of the function. The semantics prefers values from the call site, which is captured by
the $\uplus$ operation. In application, we first evaluate the expression $e_1$, then $e_2$ and finally
call the returned function. The three calls use (possibly overlapping) restrictions of the dictionary
as required by the static types.

@@

When using the usualmonadic semantics based on the reader monad, the semantics of the (\emph{abs})
rule would be modified as follows:
%
\begin{equation*}
\hspace{-0.5em}
\begin{array}{l}
  \sem{\coctx{\Gamma}{\cclrd{\emptyset}} \vdash \lambda y. e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } = 
      \lambda ((x_1, \ldots, x_n), \_) \rightarrow\\[-0.25em]
  \qquad\lambda (y, g) \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\cclrd{s}} \vdash e : \tau_2 }~((x_1, \ldots, x_n, y), g)      
\end{array}
\end{equation*}
%
Note that the declaration site dictionary
is ignored and the body is called with only the dictionary provided by the call site. This is
a consequence of the fact that monadic functions are always pure values created using monadic
\emph{unit}, which turns a function $\tau_1 \rightarrow \mtyp{\cclrd{r}}{\tau_2}$ into a monadic
computation with no side-effects $\mtyp{\cclrd{\emptyset}}{\tau_1 \rightarrow \mtyp{\cclrd{r}}{\tau_2}}$.

@@

\paragraph{Summary.}
Implicit parameters are the simplest example of a system where function abstraction does not 
``delay'' all impurities of the body. Here, the term ``delay'' refers to the fact that some 
implicit parameters may be captured (from the declaration site) at the time when the function is 
defined, but before it is executed. As discussed in Section~\ref{sec:applications-structure-lam},
this is the defining feature of \emph{coeffect} systems.

In this section, we have seen how this affects both the type system and the semantics of the 
language. In the type system, the (\emph{abs}) rule places context-requirements on both the 
declaration site and the call site. For implicit parameters, this rule introduces non-determinism
in the type-inference, because the parameters can be split arbitrarily. However, as we show in the 
next section, this is not always the case. Semantically, lambda abstraction \emph{merges} two 
parts of context (implicit parameter dictionaries) that are provided by the call site and 
declaration site.

@@

The type system and semantics for rebindable resources are essentially the same as those for
implicit parameters. Primitive requirements are introduced by the $\kvd{access}$ keyword. 
Lambda abstraction splits the requirements between declaration site (capturing remote reference) 
and call site (representing rebinding). For this reason, we do not discuss the system in details
and instead look at other uses.

@@

Sub-coeffecting and application are trivially dual to the rules for implicit parameters. We just track
supported environments using intersection as opposed to tracking required parameters using union.
However, this symmetry does not hold for lambda abstraction (\emph{abs}), which still uses \emph{union}.
This models the case when there are two ways of executing the function:
%
\begin{itemize}
\item[--] The function is represented as executable code for an call site environment
  and is executed there, possibly after it is marshalled and transferred to another machine.
\vspace{-0.5em}
\item[--] The function body is compiled for the declaration site environment; the value
  that is returned is a remote reference to the code and function calls are performed as remote invocations.
\end{itemize}
%
This example ignores important considerations -- for example, it is likely desirable to make this
difference explicit (\eg~using explicit wrapping of unevaluated expressions) and the implementation
also needs to be clarified. For a system that does this, see \eg~ML5 \cite{app-distributed-ml5}).
The key point of our brief example is that the algebraic structure of coeffect annotations may be more 
complex and use, for example, $\cap$ for application and $\cup$ for abstraction.

The second system, shown in Figure~\ref{fig:applications-flat-cross} (b) is inspired by the API
level requirements in Android. Coeffect annotations are simply numbers representing the level
($\cclrd{r}, \cclrd{s}, \cclrd{t} \in \mathbb{N}$). Levels are ordered increasingly, so we can
always require higher level (\emph{sub}). The requirement on function application (\emph{app}) is
the highest level of the levels required by the sub-expressions and the function. The system uses 
yet another variant of lambda abstraction (\emph{abs}) -- the requirements of the body are duplicated
and placed on \emph{both} the declaration site and the call site.

@@

\paragraph{Flat liveness analysis.}
In this section, we discuss a restricted form of liveness analysis. We do not track liveness of 
\emph{individual} variables, but of the \emph{entire} variable context. This is not practically
useful, but it provides an interesting insight into how flat coeffects work. A per-variable liveness
analysis can be captured using structural coeffects and is discussed in Section~\ref{sec:applications-struct-live}.
Consider the following two examples:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{constant42} = \lambda \ident{x} \rightarrow 42\\
\kvd{let}~\ident{constant} = \lambda \ident{value} \rightarrow \lambda \ident{x} \rightarrow \ident{value}
\end{array}
\end{equation*}
%
The body of the first function is just a constant $42$ and so the context of the body is marked
as \emph{dead}. The parameter (call site) of the function is not used and can also be marked as dead. 
Similarly, no variables from the declaration site are used and so they are also marked as dead.

In contrast, the body of the second function accesses a variable \ident{value} and so the body 
of the function is marked as \emph{live}. In the flat system, we do not track \emph{which} 
variable was used and so we have to mark both the call site and the declaration site as live (this will
be refined in a structural version of the system).

@@

The (\emph{abs}) rule duplicates the annotation of the body, similarly to the cross-compilation 
example in Figure~\ref{fig:applications-flat-cross}. When the body accesses any variables, it 
requires both the argument and the variables from declaration site. When it does not use any variables,
it marks both as dead. Finally, the (\emph{let}) rule annotates the composed expression with the
liveness of the expression $e_2$ -- if the context of $e_2$ is live, then it also requires variables
from $\Gamma$; if it is dead, then it does not require $\Gamma$ or $x$. 
As further discussed later in Section~?, the (\emph{let}) rule is again just a syntactic sugar for 
$(\lambda x.e_2)~e_1$. This follows from the simple observation that 
$\cclrd{r} \sqcup (\cclrd{s} \sqcap \cclrd{r}) = \cclrd{r}$.

@@

\paragraph{Examples.}
Before looking at the semantics, we consider a number of simple examples to demonstrate the
key aspects of the system. Full typing derivations are shown in Appendix~\ref{sec:appendixa-liveness}:
%
\begin{equation*}
\begin{array}{lll}
(\lambda x . 42)~y &~\hspace{1em}~&(1)\\
\ident{twoTimes}~42          &&(2)\\
(\lambda x . x)~42 &&(3)\\
\end{array}
\end{equation*}
%
In the first case, the context is dead. In ($1$), the function's parameter is dead and so the
overall context is dead, even though the argument uses a variable $y$ -- the semantics evaluates 
the function without passing it an actual argument. In the second case ($2$), the function is
a variable that needs to be obtained and so the context is live. In the last case ($3$), the 
function accesses a variable and so its declaration site is marked as requiring the context 
(\emph{abs}). This is where structural coeffect analysis would be more precise -- the system shown
here cannot capture the fact that $x$ is a bound variable.

@@

\paragraph{Summary.}
Unlike with implicit parameters, lambda abstraction for liveness analysis does not introduce 
non-determinism. It simply duplicates the context requirements. However, this still matches the
property of coeffects that impurities cannot be delayed or thunked and attached just to the 
function arrow -- we place requirements on both call site and declaration site.

@@

In function abstraction (\emph{abs}), the requirements of the body are duplicated on the declaration site
and the call site as in liveness analysis. If the body requires $\cclrd{n}$ past values, it may access
$\cclrd{n}$ values of any variables -- including those available in $\Gamma$, as well as the parameter
$x$. Finally, the (\emph{let}) rule simply adds the two requirements. This corresponds to the sequential
composition operation, but it is also a rule that we obtain by treating let-binding as a syntactic 
sugar for $(\lambda x.e_2)~e_1$.

@@

Lambda abstractions (\emph{abs}) receives two lists of the same size -- one containing values of
the variables (list of products) from the declaration site $\langle \mathbf{v}_0, \ldots, \mathbf{v}_{\cclrd{n}} \rangle$ 
and one containing the argument (list of values) provided by the call site $\langle y_0, \ldots, y_{\cclrd{n}} \rangle$.
The semantics applies the well-known \emph{zip} operation on the lists and passes the result to the
body.

@@

\paragraph{Summary.}
Despite attaching annotations to the variable context, the system for safe locking uses 
effect-style lambda abstraction. Lambda abstraction associates all requirements with the 
call site -- a lambda function created under a lock cannot access protected memory available at 
the time of creation. It will be executed later and can only access the memory available then.
This suggests that safe locking is better seen as an effect system.

@@

\paragraph{Structural liveness.} 
Recall two examples discussed earlier where the flat liveness analysis marked the whole context
as (syntactically) live, despite the fact part of it was (semantically) dead:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{constant} = \lambda y \rightarrow \lambda x \rightarrow y\\[-0.25em]
\kvd{let}~\ident{answer} = (\lambda x \rightarrow x)~42
\end{array}
\end{equation*}
%
In the first case, the variable $x$ is dead, but was marked as live. In the second example, the
declaration site of the \ident{answer} value is dead, but was marked as live. This is because in 
both of the expressions, \emph{some} variable is accessed. However, the (\emph{abs}) rule of flat
liveness has no way of determining \emph{which} variables are used by the body -- and, in particular,
whether the accessed variable is the \emph{bound} variable or some of the \emph{free} variables.

@@

\paragraph{Semantics.}
In the previous examples, we defined the semantics -- somewhat informally -- using a simple 
$\lambda$-calculus language to encode the model. More formally, this could be a Cartesian-closed 
category. In that model, we can reuse variables arbitrarily and so it is not
a good fit for modelling bounded reuse. Girard et al. \cite{logic-bounded} model their bounded
linear logic in an (ordinary) linear logic where variables can be used at most once.

@@

\paragraph{Example.} As an example, consider a function $\lambda x.\kvd{prev}~(y+x)$ applied to an argument
$\kvd{prev}~(\kvd{prev}~y)$. The body of the function accesses the past value of two variables, one free
and one bound. The (\emph{abs}) rule splits the annotations between the declaration site and call site
of the function:
%
\begin{equation*}
\tyrule{abs}
  {\coctx{y\!:\!\mathbb{Z}, x\!:\!\mathbb{Z}}{\alift{1, 1}} \vdash \kvd{prev}~(y+x) : \mathbb{Z} }
  {\coctx{y\!:\!\mathbb{Z}}{\alift{1}} \vdash \lambda x . \kvd{prev}~(y+x) : \mathbb{Z} \xrightarrow{\cclrd{1}} \mathbb{Z} }
\end{equation*}
%
The expression always requires the previous value of $y$ and adds it to a previous value of the 
parameter $x$. Evaluating the value of the argument $\kvd{prev}~(\kvd{prev}~y)$ requires two past 
values of $y$ and so the overall requirement for the (free) variable $y$ is $3$ past values. In 
order to use the contraction rule, we rename $y$ to $y'$ in the argument:
%
\begin{equation*}
\inference
  { \coctx{y\!:\!\mathbb{Z}}{\alift{1} } \vdash \lambda x.~(\ldots) : \mathbb{Z} \xrightarrow{\cclrd{1}} \mathbb{Z} &
    \coctx{x\!:\!\mathbb{Z}}{\alift{2}} \vdash (\kvd{prev}~(\kvd{prev}~y') : \mathbb{Z} }
{\inference
  { \coctx{y\!:\!\mathbb{Z}, y'\!:\!\mathbb{Z}}{\alift{1,3}} \vdash (\lambda x.\kvd{prev}~(y+x))~(\kvd{prev}~(\kvd{prev}~y')) : \mathbb{Z} }
  { \coctx{y\!:\!\mathbb{Z}}{\alift{3}} \vdash (\lambda x.\kvd{prev}~(y+x))~(\kvd{prev}~(\kvd{prev}~y)) : \mathbb{Z} } }
\end{equation*}
%
The derivation uses (\emph{app}) to get requirements $\alift{1,3}$ and then (\emph{contr}) to take 
the maximum, showing three past values are sufficient.
